Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-03-08 supplement.tex
edits to figure captions and supp. cleanup

\citet{Ragsdale2019-nt} showed how to compute the analogous multi-population
set of LD statistics, and we refer readers there for details on their
definitions, computation, and interpretations. In short, we obtain expectations
of $D^2$ in each population, the cross-population product $D_iD_j$ (where $i$
and $j$ index populations), as well as those additional statistics $Dz$ and
$\pi_2$ taken over different combinations of population indexing. That is
$Dz_{i, j, k} = D_i(1-2p_{A,j})(1-2p_{B,k})$ and
$\pi_{2;i,j,k,l}=p_{A,i}(1-p_{A,j})p_{B,k}(1-2p_{B,l})$. We consider statistics
normalized by $\pi_2$ in a reference population (throughout, we use the Mende
\comment{Why use the Mende to normalize?}
$\pi_2$), which removes any dependence on the mutation rate. Thus, statistics
take the form $\sigma_{d;i,j}^2 = \frac{\E[D^2_{i,j}]}{\E[\pi_2]}$,
$\sigma_{Dz;i,j,k}=\frac{\E[Dz_{i,j,k}]}{\E[\pi_2]}$, and so on.
Multi-population pairwise diversity statistics $\E[H_i]=\E[2p_i(1-p_i)]$ and
$\E[H_{i,j}] = \E[p_i(1-p_j) + p_j(1-p_i)]$ were also normalized by $H$ in the
Mende, so that all pairwise diversity measures are relative to the reference
population.

@@

Here we focus on four parameterizations of early human history: (1) a “single
origin” with recent expansion, (2) primary and second stem populations with
“continuous migration”, (3) two stem populations that do not exchange migrants
before merging to form regional African populations, referred to as a “merger
without stem migration”, and (4) two stem populations that are allowed to
exchange migrants before merging to form regional African populations, referred
to as a “merger with stem migration”. Given the best fit parameters for each
model
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
we used \msprime version 1.0.4 \citep{Kelleher2016-lw,Baumdicker2021-mj} to
simulate sampled individual genomes to match the data used in inference (see
simulation details below). Using these simulations, we compared additional
statistics and inferences between simulated and real data that were not used in
the original optimization.
\comment{Describe in more detail}
These comparisons serve as independent validations
of our inferred demographic models by assessing how well those models fit
aspects of the data that have been used in other studies as evidence of
population structure or archaic hominin admixture from an unidentified hominin species
within Africa.

@@

\begin{table}[ht]
\caption{
    \label{tab:supp-continuous-migration-no-mig}
    \textbf{Best-fit parameters from the Continuous-Migration model with isolation
    between stems.}
    Inferred values are scaled to physical units assuming a generation time of
    29 years. This model gave a log-likelihood of -126,644. This model allows for
    continuous migration between Stem 2 and modern populations, but disallows
    migration between Stem 2 and Stem 1, which later splits into branches leading
    to South, East, and West African populations.
}
\centering
\begin{tabular}[t]{rP{8cm}cc}
    \toprule
    Parameter & Description & Value & Std. err.\\
    \midrule
    $N_e$ & Ancestral effective population size & 10543 & 460 \\
    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal and Nama splits & 15297 & 1227 \\
    $N_{stem2}$ & Size of stem 2 lineage & 9156 & 573 \\
    $N_{Nama_0}$ & Initial Nama size & 11329 & 423 \\
    $N_{Nama_F}$ & Final Nama size & 219 & 11 \\
    $N_{MSL_0}$ & Initial Mende size & 11187 & 644 \\
    $N_{MSL_F}$ & Final Mende size & 24056 & 945 \\
    $N_{EA}$ & Size of East African branch & 7185 & 293 \\
    $N_{Gumuz_F}$ & Final Gumuz size & 3746 & 184 \\
    $N_{EP}$ & East African agriculturist size & 13187 & 645 \\
    $N_{GBR_0}$ & Initial British size & 937 & 41 \\
    $N_{GBR_F}$ & Final British size & 11478 & 420 \\
    $N_{Neand}$ & Neanderthal size & 2285 & 113 \\
    $T_{stems}$ & Stem split time (years) & 459720 & 21704 \\
    $T_{Nama}$ & Nama split time (years) & 139725 & 3250 \\
    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $1.93\times10^{-5}$ & $0.360\times10^{-5}$ \\
    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.43\times10^{-5}$ & $0.309\times10^{-5}$ \\
    $m_{MSL-EA}$ & Mende--East Africa migration rate & $20.8\times10^{-5}$ & $0.97\times10^{-5}$ \\
    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.46\times10^{-5}$ & $0.171\times10^{-5}$ \\
    $m_{EA-EA}$ & Intra-East Africa migration rate & $35.4\times10^{-5}$ & $1.85\times10^{-5}$ \\
    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12 ka ($1-f$ from Gumuz) & 0.649 & 0.0038 \\
    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2 ka & 0.268 & 0.0048 \\
    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.153 & 0.0022 \\
    $m_{stems}$ & Stem 1--stem 2 migration rate & $0$ & fixed \\
    $m_{stem2-Nama}$ & Stem 2--Nama migration rate & $4.33\times10^{-5}$ & $0.891\times10^{-5}$ \\
    $m_{stem2-MSL}$ & Stem 2--Mende migration rate & $11.3\times10^{-5}$ & $1.83\times10^{-5}$ \\
    $m_{stem2-EA}$ & Stem 2--East Africa migration rate & $2.77\times10^{-5}$ & $0.699\times10^{-5}$ \\
    \bottomrule
\end{tabular}
\end{table}

@@

\clearpage
\comment{30 is a lot of supp figures... can these be condensed to be a bit less daunting to readers?}

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.65\textwidth]{figures/supp-possible-models.png}
    \caption{
        \textbf{Proposed models of early human history in Africa.}
        \comment{Lovely figure, may even want to move to main text}
    }
    \label{fig:supp-possible-models}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-omni-hapmap-comparison.pdf}
    \caption{
        \textbf{Comparison of LD statistics parsed using two recombination maps}
        The recombination map determines genetic distances between pairs of SNPs,
        so using different maps could result in systematic biases in parsed
        statistics if the maps differ significantly. Here, the solid lines are
        statistics parsed using the HapMapII recombination map, and dashed lines
        are using the OMNI-YRI map. Shading represents 95\% confidence intervals
        from the OMNI map. These two maps results in LD statistics that are
        indistinguishable from one another. Notations for each statistic are
        described in section~\ref{sec:supp-ld-stats} and indexed by populations
        (0: Nama, 1: Mende, 2: Gumuz, 3: Amhara/Oromo, 4: British).
    }
    \label{fig:supp-omni-hapmap-comparison}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-continuous-migration-fits.pdf}
    \caption{
        \textbf{Continuous-migration model fit to LD statistics.}
        Predicted vs observed two-locus statistics as a function of genetic
        distance for the continuous-migration model. Each panel represents a different
        set of two-locus statistics. Solid lines represent estimates using the
        continuous-migration model. Dashed lines represent observed statistics.
        Notation and indexing of statistics are described in the
        Fig.~\ref{fig:supp-omni-hapmap-comparison} caption.
    }
    \label{fig:supp-continuous-migration-fits}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-fsts-all-models.pdf}
    \caption{
        \textbf{Predicted \bm{$F_{ST}$} between contemporaneous populations over time.}
        All models (A: single origin, B: continuous migration, C: merger without stem
        migration, D: merger with stem migration) match observed $F_{ST}$ between
        sampled populations. The continuous-migration model predicts that human stem
        populations remain genetically similar, while the merger with and without
        stem migration models predict a period of increased $F_{ST}$ between ancestors
        of modern humans. This is largely due to the very small inferred $N_e$ in one of
        the stem branches, which leads to a rapid increase in differentiation due to
        drift.
    }
    \label{fig:supp-FST-panels}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-h12s-all-models.pdf}
    \caption{
        \textbf{Predicted pairwise differentiation between contemporaneous
        populations.} $H_{i,j}$ is the predicted pairwise differentiaion
        between two populations per base-pair, using a mutation rate estimated
        from matching predicted to observed pairwise diversity from the
        intergenic data used in the fits (see Sec.~\ref{sec:supp-mutation}).
        The single-origin model (A) shows larger deviations between observed
        and expected differentiation than the models with early human structure
        (B: continuous migration, C: merger without stem migration, D: merger
        with stem migration). In the merger models (C and D), while $F_{ST}$ is
        large between early stems due to the sharp bottleneck in one of those
        branches, the pairwise differences are comparable to the
        continuous-migration model. Stem 1E and Stem 1S have low
        differentiation, as the bottleneck in Stem 1 (which they both split
        from) sharply reduces diversity by the time of the split. This has the
        effect of large $F_{ST}$ but small $H_{i,j}$.
    }
    \label{fig:supp-h12-panels}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-f4s-single-origin.pdf}
    \caption{
        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
        and ancient branches from the single-origin model.}
        $f_4(A, B; X, Y)/f_2(A, B)f_2(X, Y)$ is interpreted as the proportion of the
        amount of drift between sampled populations $A$ and $B$ that aligns with
        drift between populations $X$ and $Y$, sampled in the past
        (see Section~\ref{sec:f4}). In the single-origin model, nearly all
        differentiation between the Nama, Mende, and Gumuz populations arises
        after the divergence events, so normalized $f_4$ decays to zero by the
        time of those divergences.
    }
    \label{fig:supp-f4s-single-origin}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-f4s-continuous-migration.pdf}
    \caption{
        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
        and ancient branches from the continuous-migration model.}
        Despite population structure inferred to have extended up to 1Ma
        into the past, differentiation between modern African populations
        only traces back to differentiation between ancestral populations
        $\sim$100-200ka, indicating that while stem populations were structured
        (Fig.~\ref{fig:supp-FST-panels}), differentiation due to that
        ancient structure does not contribute to differention between modern
        populations.
    }
    \label{fig:supp-f4s-continuous-migration}
\end{figure}

>>>>

\citet{Ragsdale2019-nt} showed how to compute the analogous multi-population
set of LD statistics, and we refer readers there for details on their
definitions, computation, and interpretations. In short, we obtain expectations
of $D^2$ in each population, the cross-population product $D_iD_j$ (where $i$
and $j$ index populations), as well as those additional statistics $Dz$ and
$\pi_2$ taken over different combinations of population indexing. That is
$Dz_{i, j, k} = D_i(1-2p_{A,j})(1-2p_{B,k})$ and
$\pi_{2;i,j,k,l}=p_{A,i}(1-p_{A,j})p_{B,k}(1-2p_{B,l})$. We consider statistics
normalized by $\pi_2$ in an arbitrary reference population (throughout, we use the Mende
$\pi_2$), which removes any dependence on the mutation rate. Thus, statistics
take the form $\sigma_{d;i,j}^2 = \frac{\E[D^2_{i,j}]}{\E[\pi_2]}$,
$\sigma_{Dz;i,j,k}=\frac{\E[Dz_{i,j,k}]}{\E[\pi_2]}$, and so on.
Multi-population pairwise diversity statistics $\E[H_i]=\E[2p_i(1-p_i)]$ and
$\E[H_{i,j}] = \E[p_i(1-p_j) + p_j(1-p_i)]$ were also normalized by $H$ in the
Mende, so that all pairwise diversity measures are relative to the reference
population.

@@

Here we focus on four parameterizations of early human history: (1) a “single
origin” with recent expansion, (2) primary and second stem populations with
“continuous migration”, (3) two stem populations that do not exchange migrants
before merging to form regional African populations, referred to as a “merger
without stem migration”, and (4) two stem populations that are allowed to
exchange migrants before merging to form regional African populations, referred
to as a “merger with stem migration”. Given the best fit parameters for each
model
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
we used \msprime version 1.0.4 \citep{Kelleher2016-lw,Baumdicker2021-mj} to
simulate sampled individual genomes to match the data used in inference (see
simulation details below). We used these simulations to compare our model predictions to 
predictions based on the conditional SFS (cSFS) and on relate curves (see below).

These comparisons serve as independent validations
of our inferred demographic models by assessing how well those models fit
aspects of the data that have been used in other studies as evidence of
population structure or archaic hominin admixture from an unidentified hominin species
within Africa.

@@

\begin{table}[ht]
\caption{
    \label{tab:supp-continuous-migration-no-mig}
    \textbf{Best-fit parameters from the Continuous-Migration model without stem migration.}
    Inferred values are scaled to physical units assuming a generation time of
    29 years. This model gave a log-likelihood of -126,644. This model allows for
    continuous migration between Stem 2 and modern populations, but disallows
    migration between Stem 2 and Stem 1, which later splits into branches leading
    to South, East, and West African populations.
}
\centering
\begin{tabular}[t]{rP{8cm}cc}
    \toprule
    Parameter & Description & Value & Std. err.\\
    \midrule
    $N_e$ & Ancestral effective population size & 10543 & 460 \\
    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal and Nama splits & 15297 & 1227 \\
    $N_{stem2}$ & Size of stem 2 lineage & 9156 & 573 \\
    $N_{Nama_0}$ & Initial Nama size & 11329 & 423 \\
    $N_{Nama_F}$ & Final Nama size & 219 & 11 \\
    $N_{MSL_0}$ & Initial Mende size & 11187 & 644 \\
    $N_{MSL_F}$ & Final Mende size & 24056 & 945 \\
    $N_{EA}$ & Size of East African branch & 7185 & 293 \\
    $N_{Gumuz_F}$ & Final Gumuz size & 3746 & 184 \\
    $N_{EP}$ & East African agriculturist size & 13187 & 645 \\
    $N_{GBR_0}$ & Initial British size & 937 & 41 \\
    $N_{GBR_F}$ & Final British size & 11478 & 420 \\
    $N_{Neand}$ & Neanderthal size & 2285 & 113 \\
    $T_{stems}$ & Stem split time (years) & 459720 & 21704 \\
    $T_{Nama}$ & Nama split time (years) & 139725 & 3250 \\
    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $1.93\times10^{-5}$ & $0.360\times10^{-5}$ \\
    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.43\times10^{-5}$ & $0.309\times10^{-5}$ \\
    $m_{MSL-EA}$ & Mende--East Africa migration rate & $20.8\times10^{-5}$ & $0.97\times10^{-5}$ \\
    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.46\times10^{-5}$ & $0.171\times10^{-5}$ \\
    $m_{EA-EA}$ & Intra-East Africa migration rate & $35.4\times10^{-5}$ & $1.85\times10^{-5}$ \\
    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12 ka ($1-f$ from Gumuz) & 0.649 & 0.0038 \\
    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2 ka & 0.268 & 0.0048 \\
    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.153 & 0.0022 \\
    $m_{stems}$ & Stem 1--stem 2 migration rate & $0$ & fixed \\
    $m_{stem2-Nama}$ & Stem 2--Nama migration rate & $4.33\times10^{-5}$ & $0.891\times10^{-5}$ \\
    $m_{stem2-MSL}$ & Stem 2--Mende migration rate & $11.3\times10^{-5}$ & $1.83\times10^{-5}$ \\
    $m_{stem2-EA}$ & Stem 2--East Africa migration rate & $2.77\times10^{-5}$ & $0.699\times10^{-5}$ \\
    \bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.65\textwidth]{figures/supp-possible-models.png}
    \caption{
        \textbf{Models of early human history in Africa} seeking to reconcile genetic and paleoanthropological evidence \cite{Henn2018-rf}.
         \comment{Lovely figure, may even want to move to main text}
    }
    \label{fig:supp-possible-models}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-omni-hapmap-comparison.pdf}
    \caption{\sgcomment{$\pi$ in figure are not rendered properly}
        \textbf{Comparison of LD statistics parsed using two recombination maps}
        The recombination map determines genetic distances between pairs of SNPs,
        so using different maps could result in systematic biases in parsed
        statistics if the maps differ significantly. Here, the solid lines are
        statistics parsed using the HapMapII recombination map, and dashed lines
        are using the OMNI-YRI map. Shading represents 95\% confidence intervals
        from the OMNI map. These two maps results in LD statistics that are
        indistinguishable from one another. Notations for each statistic are
        described in section~\ref{sec:supp-ld-stats} and indexed by populations
        (0: Nama, 1: Mende, 2: Gumuz, 3: Amhara/Oromo, 4: British).
    }
    \label{fig:supp-omni-hapmap-comparison}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-continuous-migration-fits.pdf}
    \caption{
        \textbf{Continuous-migration model fit to LD statistics.}
        Predicted vs observed two-locus statistics as a function of genetic
        distance for the continuous-migration model \sgcomment{with or without migration?}. Each panel represents a different
        set of two-locus statistics. Solid lines represent estimates using the
        continuous-migration model. Dashed lines represent observed statistics.
        Notation and indexing of statistics are described in the
        Fig.~\ref{fig:supp-omni-hapmap-comparison} caption.
    }
    \label{fig:supp-continuous-migration-fits}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-fsts-all-models.pdf}
    \caption{\sgcomment{Need ABCD in figure!}
        \textbf{Predicted \bm{$F_{ST}$} between contemporaneous populations over time.}
        All models (A: single origin, B: continuous migration, C: merger without stem
        migration, D: merger with stem migration) match observed $F_{ST}$ between
        sampled populations. The continuous-migration model predicts that human stem
        populations remain genetically similar, while the merger with and without
        stem migration models predict a period of increased $F_{ST}$ between ancestors
        of modern humans. This is largely due to the very small inferred $N_e$ in one of
        the stem branches, which leads to a rapid increase in differentiation due to
        drift.
    }
    \label{fig:supp-FST-panels}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-h12s-all-models.pdf}
    \caption{\sgcomment{can we sync with text colors? Need ABCD in figure}
        \textbf{Predicted pairwise differentiation between contemporaneous
        populations.} $H_{i,j}$ is the predicted pairwise differentiation
        between two populations per base-pair, using a mutation rate estimated
        from matching predicted to observed pairwise diversity from the
        intergenic data used in the fits (see Sec.~\ref{sec:supp-mutation}).
        The single-origin model (A) shows larger deviations between observed
        and expected differentiation than the models with early human structure
        (B: continuous migration, C: merger without stem migration, D: merger
        with stem migration). In the merger models (C and D), while $F_{ST}$ is
        large between early stems due to the sharp bottleneck in one of those
        branches, the pairwise differences are comparable to the
        continuous-migration model. Stem 1E and Stem 1S have low
        differentiation, as the bottleneck in Stem 1 (which they both split
        from) sharply reduces diversity by the time of the split. This has the
        effect of large $F_{ST}$ but small $H_{i,j}$.
    }
    \label{fig:supp-h12-panels}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-f4s-single-origin.pdf}
    \caption{
        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
        and ancient branches from the single-origin model.}
        $f_4^2(A, B; X, Y)/f_2(A, B)f_2(X, Y)$ is interpreted as the proportion of the
        amount of drift between sampled populations $A$ and $B$ that aligns with
        drift between populations $X$ and $Y$, sampled in the past
        (see Section~\ref{sec:f4}). In the single-origin model, nearly all
        differentiation between the Nama, Mende, and Gumuz populations arises
        after the divergence events, so normalized $f_4$ decays to zero by the
        time of those divergences.
    }
    \label{fig:supp-f4s-single-origin}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-f4s-continuous-migration.pdf}
    \caption{
        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
        and ancient branches from the continuous-migration model.}
        Despite population structure inferred to have extended up to 1Ma
        into the past, differentiation between modern African populations
        only traces back to differentiation between ancestral populations
        $\sim$100-200ka, indicating that while stem populations were structured
        (Fig.~\ref{fig:supp-FST-panels}), differentiation due to that
        ancient structure contributes little to differentiation between modern
        populations.
    }
    \label{fig:supp-f4s-continuous-migration}
\end{figure}
